Learning at the Racetrack: Data-driven Methods to Improve Racing Performance over Multiple Laps

### 2019-09-03 journal/TCST2019/manuscript.tex
modifications to tcst based on reviewer feedback

\begin{abstract}
Autonomous vehicles will generate tremendous data from the variety of sensors they employ to track the surrounding environment. This data is inherently valuable, as it gives algorithm designers the potential to leverage prior experience in order to improve driving performance over time. This paper uses the lens of autonomous racing to provide an example of how data from previous iterations of driving can be used to improve quantitative metrics of performance. Two complementary algorithms are demonstrated in this paper. The first type uses iterative learning control (ILC) to simultaneously improve lateral and longitudinal tracking of the desired racing trajectory from lap-to-lap, while the second type of algorithm is focused on altering the trajectory itself using a search method. When driven experimentally at the limits of handling, the result is a reduction in lap time of nearly 1.4 seconds, a major improvement.

@@

\subsection{Problem Description and Proposed Solution}
Several methods exist in the literature to generate autonomous racing trajectories (~\cite{casanova},\cite{kelly},\cite{theodosis},\cite{kapaniadscc}), which typically comprise of a longitudinal speed profile $U_x(s)$ and a desired curvature profile $\kappa(s)$ as a function of distance along the track. These trajectories can then be tracked using one of several path following controllers, such as (\cite{mickcop},\cite{fricStudy},\cite{mickgeneral},\cite{shladover},\cite{nagai}).

Using the trajectory generated by \cite{kapaniadscc} and the trajectory following controller presented by \cite{kapania}, the
authors generated a test run of an autonomous race vehicle (see Fig.\ref{fig:carpic}(a)) at the limits of handling on the Thunderhill Raceway course in Willows, CA (Fig.~\ref{fig:carpic}(b)). The results are shown in Fig. \ref{fig:expErrors}.

\begin{figure}[h]
\centering
\includegraphics[width=3.5in]{figures/pic1.png}
\caption{(a) ``Shelley" - Stanford's autonomous self-driving race vehicle. (b) Top-down view of Thunderhill Raceway course. }
\label{fig:carpic}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=3.3in]{figures/expErrors.eps}
\caption{Controller tracking performance and tire slip norm on a test run at the limits of handling ($\mu =$ 0.94).(a) Desired
vs actual speed of the vehicle. (b) Lateral tracking error and (c) tire slip norm as a function of distance along the path.}
\label{fig:expErrors}
\end{figure}

@@

Where $\sigma$ and $\alpha$ are the longitudinal and lateral tire slip for a given tire, and $\sigma_p$ and $\alpha_p$ are empirically
determined \textit{peak slip} values resulting in maximum longitudinal and lateral tire force generation. As a result, $\zeta < 1 $ corresponds to the tires having excess force generation capacity, while $\zeta > 1$ corresponds to tire saturation.

@@

Iterative learning control (ILC) algorithms can help an autonomous vehicle follow a desired trajectory more precisely over 
several laps of driving. However, ILC algorithms do not alter the desired trajectory itself, only the input signals that attempt
to track the trajectory. This will not be sufficient at the limits of handling. Consider Fig.~{\ref{fig:expErrors}, where the desired speed profile was generated assuming a tire-road friction value of $\mu = 0.94$. While this is a reasonable assumption overall, there are several parts on the track where the vehicle exceeds the available friction and begins to understeer. Region \circled{2} is one example. The tire slip norm $\zeta$ climbs above one, and as a result, the vehicle begins
to slide off the track, resulting in the large negative tracking error spike in Fig.~{\ref{fig:expErrors}(b). The vehicle's stability algorithms must kick in and slow the vehicle down, which results in the speed tracking error seen in Fig.~{\ref{fig:expErrors}(a). In this situation, ILC will be unable to achieve better speed tracking and path tracking performance. The tires are saturated and the car is slowly beginning to careen off the track. Simply steering more
on the next lap will not achieve better tracking performance due to the saturation of the steering actuator. In order to recover, the vehicle must deviate from the planned trajectory, either by slowing down or by taking a wider radius turn.

Some trajectory modification algorithms previously investigated in
the literature have focused on modifying the \textit{lateral} trajectory by altering the curvature profile as the vehicle understeers or oversteers. For example, Theodosis and Gerdes presented an algorithm to gradually widen the radius of a turn in response to a detected understeer \cite{paulthesis}. The algorithm was validated
experimentally, but assumed sufficient availability of road width. Klomp and Gordon also developed a strategy for recovering from
vehicle understeer by solving for an optimal emergency braking profile to minimize deviation from the desired path \cite{klomp}. On the other hand, Brown et al.~\cite{brown2017safe} presented a model predictive control (MPC) approach that generally aimed to follow a planned vehicle trajectory at the limits. However, if the vehicle was at risk of understeering or oversteering, the MPC algorithm would deviate laterally from the planned trajectory in order to maintain stability of the vehicle without driving off the road.

Another approach to trajectory modification is Learning Model Predictive Control (LMPC), developed by Rosolia et al. \cite{rosolia2017autonomous} for the purpose of autonomous racing. In LMPC, data from every lap of racing is used to build invariant sets and approximate the value function for the generally non-convex lap time minimization problem. In practice, this enables the controller to explore different areas of the state space on each successive lap, modifying the desired trajectory on each lap in order to continuously improve the lap time.

@@

Due to the repetitive nature of automotive racing, iterative learning control techniques are a promising method to gradually
eliminate \textit{trajectory tracking} errors . For this application of ILC, the repetitive trials are laps of racing, and the reference trajectory is the optimal speed profile and curvature developed in \cite{kapaniadscc}. This paper will present adaptations of two established ILC methods 
(proportional-derivative and quadratically optimal) for use in the Audi TTS racing system. Section \ref{sec:ch4dsm} presents
both coupled and decoupled models of the vehicle lateral and longitudinal dynamics. These models are converted from state-space representations to \textit{lifted domain}
representations required for iterative learning control, and the proportional-derivative and quadratically optimal ILC algorithms are presented in \S \ref{sec:pdcontroller} and \S \ref{sec:qopt}. Section \ref{sec:simres} presents simulated results of the ILC algorithms for a sample vehicle trajectory, and finally,
experimental results showing a gradual reduction of trajectory-following errors is presented in \S \ref{sec:ch4ExpRes}.

For situations where the vehicle underperforms due to excessive oversteer or understeer, an A* search algorithm is presented
that iteratively modifies portions of the \textit{velocity} profile to be more conservative if a stability violation is encountered on a prior lap. This is accomplished by generalizing the speed profile such that each part of the track can be driven with a learned  value of friction $\mu$ and therefore a different maximum
acceleration. The algorithm also modifies the velocity profile to be more aggressive if the tires are not being actuated at the limits. In this way, the algorithm is focused on learning the time-optimal friction profile $\mu^\star(s)$ by searching through datasets obtained over multiple laps.

@@

The more difficult task is expressing the dynamics of the two subsystems using an affine model, given the tendency for the vehicle tires to saturate at the limits. This paper will model the dynamics with a bicycle model with single longitudinal input $F_x$ and front steer input $\delta$, shown in Fig.\ref{fig:bikemodel}.

@@

Nonlinear, coupled equations of motion are provided in (\ref{eq:fullNL})-(\ref{eq:fullNL2}):

\begin{align}
\label{eq:fullNL}
	\dfrac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
	\dfrac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{F_x}{m} - \frac{F_\mathrm{yf}(\alpha_\mathrm{f},F_x)\delta}{m}\\
	\dfrac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)}-r \\
	\dfrac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
	\dfrac{d\Delta\Psi}{dt} &= r \label{eq:fullNL2}
\end{align}
The system dynamics presented in (\ref{eq:fullNL}) have five states, yaw rate $r$, vehicle sideslip $\beta$, lateral path tracking error $e$, heading error $\Delta \Psi$, and speed tracking error $v$, defined by:

\begin{equation}
v = U_x - U^\mathrm{des}_x
\end{equation}

The potential for coupling between the subsystems is apparent from  (\ref{eq:fullNL})-(\ref{eq:fullNL2}), not only directly from the state equations but also due to the complex nature of tire force generation at the handling limits. As shown in Fig.~\ref{fig:coupledTires}, as longitudinal
force $F_x$ is distributed across the tires, the available lateral force decreases. At the limits of handling, this
\textit{derating} of the lateral force may become significant. As a result, we model lateral tire force generation $F_y$ as a function
of both lateral tire slip $\alpha$ and the applied longitudinal force $F_x$, using a modified form of the Fiala tire model \cite{rami}. The front and rear tire slip are themselves functions of the vehicle state and are given by:

\begin{subequations}
\begin{align}
	\alpha_\mathrm{f} &= \beta + \frac{ar}{U_x} - \delta\\
	\alpha_\mathrm{r} &= \beta - \frac{br}{U_x}
\end{align}
\end{subequations}

\begin{figure}[tb]
\centering
\includegraphics[width=2.8in]{figures/tirecurves.eps}
\caption{Lateral tire force curve as a function of longitudinal force $F_x$ and lateral tire slip $\alpha$.}
\label{fig:coupledTires}
\end{figure}

The next step is to break up the control inputs into the closed-loop feedback term and the \textit{learned} component that is
modified on every lap by the ILC algorithm:
\begin{align}
\label{eqn:splitTerms}
F_x &= F^{\mathrm{FB}}_x + F^L_x\\
    &= -K_xv + F^L_x\\
\delta &= \delta_{\mathrm{FB}} + \delta_L\\
\delta &= -k_p(e + x_\mathrm{LA}\Delta\Psi) + \delta_L
\end{align}
where $K_x$ is the proportional speed tracking gain and $k_p$ and $x_\mathrm{LA}$ are the lookahead gains presented in \cite{kapania}.

The closed loop system dynamics are now given by: 
\begin{align}
\label{eq:fullNLCL}	
	\frac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
	\frac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{-K_xv + F^L_x}{m} \\
	& - \frac{F_\mathrm{yf}(\alpha_F,F_x)(k_p(e + x_{LA}\Delta\Psi) + \delta_L)}{m}\\
	\frac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)} - r \\
	\frac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
	\frac{d\Delta\Psi}{dt} &= r \label{eq:fullNLCL2}	
\end{align}

The nonlinear closed-loop dynamics must be converted into an affine, discrete-time dynamical system to apply conventional iterative learning
control algorithms. In our case, we have two system outputs ($y = [e \hspace{2mm} v]^T$) that are measured and two input signals to learn ($u = [\delta_L \hspace{2mm} F^L_x]^T$). 
Since we run the iterative learning control algorithm after seeing a trial of data, we can approximate
the dynamics in (\ref{eq:fullNLCL})-(\ref{eq:fullNLCL2}) by linearizing about the observed states and inputs $x_o, u_o$ from the first lap. The affine model is therefore given by: 
 \begin{align}
 \label{eq:ssMAT}
 \begin{bmatrix} \dot{e} \\ \dot{\beta} \\ \dot{r} \\ \dot{\Delta\Psi} \\ \dot{v}\end{bmatrix} &= \begin{bmatrix} A(t) \end{bmatrix} \begin{bmatrix} e \\ \beta \\ r \\ \Delta\Psi \\ v \end{bmatrix} + \begin{bmatrix} B(t) \end{bmatrix} \begin{bmatrix} \delta_L \\ F^L_x \end{bmatrix} + d(t) \\
 \begin{bmatrix} \dot{e} \\ \dot{v} \end{bmatrix}  &= \begin{bmatrix} 1 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 1 \end{bmatrix} x
 \end{align}

@@

 While this multiple-input, multiple-output (MIMO) model captures the coupled behavior of the longitudinal and lateral inputs, it is tedious to compute,
 either numerically or analytically. If we make the assumption that the longitudinal and lateral inputs are decoupled and make small angle approximations, we obtain the following linear state equations for the lateral dynamics: 
\begin{subequations}
\label{eq:bmC4}
\begin{align}
	\dot{\beta} &= \frac{F_\mathrm{yf}+F_\mathrm{yr}}{mU_x} - r \qquad \dot{r} = \frac{aF_\mathrm{yf} - bF_\mathrm{yr}}{I_z} \label{bm1} \\
	\dot{e} &= U_x (\beta + \Delta\Psi) \qquad \Delta\dot{\Psi} = r - U_x\kappa \label{eq:bm2} 
\end{align}
\end{subequations} 
and the following first order equation for the longitudinal dynamics, assuming a simple point mass model with proportional
speed tracking feedback:
\begin{equation}
\dot{v} = \frac{-K_xv + F^L_x}{m}
\end{equation}
The resulting state matrices $A(t)$ and $B(t)$ for (\ref{eq:ssMAT}) are then simply block diagonal matrices consisting of the
linearized lateral dynamics and the first order longitudinal dynamics:

\begin{align}
\label{eqn:C4dyn}
 &A(t)  = \\ \nonumber
  &\left[\begin{smallmatrix}
  0 & U_x(t) & 0 & U_x(t) & 0\\ 
  0 & 0 & 1 & 0 & 0 \\ 
  \frac{-ak_\mathrm{p} \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}}  & \frac{-ak_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}}  & \frac{-a^2\tilde{C}_\mathrm{f}(t)-b^2\tilde{C}_\mathrm{r}(t)}{U_x(t)I_\mathrm{z}} & \frac{b\tilde{C}_\mathrm{r}(t) - a\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 \\
  \frac{-k_\mathrm{p}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{-k_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{b\tilde{C}_\mathrm{r}(t)-a\tilde{C}_\mathrm{f}(t)}{mU_x(t)^2}-1 & \frac{-\tilde{C}_\mathrm{f}(t) - \tilde{C}_\mathrm{r}(t)}{mU_x(t)} & 0 \\
  0 & 0 & 0 & 0 & -K_x
  \end{smallmatrix}\right] \\
&B(t) =\begin{bmatrix} 0 & 0 \\	 	0 & 0 		\\ 		\frac{a \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 		\\ 		\frac{\tilde{C}_\mathrm{f}(t)}{mU_x(t)} & 0 		\\ 0 & 1 \end{bmatrix} 
\end{align}
The affine term $d(t)$ is given by: 
\begin{align}
d(t) = \left[\begin{matrix} 0 \\
               -\kappa(t) U_x(t) \\ 
			    \frac{a\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) - b\tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + a\tilde{F}_\mathrm{yf}(t) - b\tilde{F}_\mathrm{yr}(t)}{I_z}\\
				\frac{\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) + \tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + \tilde{F}_\mathrm{yf}(t) + \tilde{F}_\mathrm{yr}(t)}{mU_x(t)}\\
				0 \\
				0
				\end{matrix}\right]
\end{align}

@@

\subsection{Lifted Domain Representation and ILC \newline Problem Statement}
\label{sec:liftedD}
Whether the coupled (\ref{eq:C4nl}) or decoupled (\ref{eqn:C4dyn}) dynamics are assumed, the
final modeling step is to apply standard discretization techniques to obtain dynamics in the following form:
\begin{align}
\label{C4ad}
x_{k+1} &= A_kx_k + B_ku_k + d_k \\
y_{k}   &= Cx_k
\end{align}
For a given lap of racing $j$, sensor measurements provide $N$ observations of both the lateral path deviation $e$ and longitudinal
speed tracking error $v$. These measurements can be stacked into a 2$N \times 1$ array: 
\begin{equation}
\mathbf{e}_j = \begin{bmatrix} e_1 & \hdots & e_N & v_1 & \hdots & v_N \end{bmatrix}^T
\end{equation} 
These measurement errors are related to the learned control inputs $\delta^L$ and $F^L_x$ as follows: 
\begin{align}
\label{eq:liftedDomain}
\mathbf{e}_j &= P\mathbf{u}^L_j + \mathbf{w} \\
\mathbf{u}^L_j &= \begin{bmatrix} \delta^L_1 & \hdots & \delta^L_N & F^L_{x1} & \hdots & F^L_{xN} \end{bmatrix}^T
\end{align}
The system dynamics modeled in the previous section are represented by the \textit{lifted-domain} dynamics matrix $P$,
which is $\mathrm{2}N \times \mathrm{2}N$ and given by:
\begin{equation}
\label{eq:C4bm}
P=\left[
\begin{array}{c|c}
P_{e\delta} & P_{eF} \\ \hline
P_{v\delta} & P_{vF} 
\end{array}\right]
\end{equation}
Where each submatrix in (\ref{eq:C4bm}) is $N \times N$ and represents the lifted-domain dynamics from a given input to a given output. 
Individual terms of the sub-matrices are given by:

\begin{equation}
\label{eq:how2Lift}
p_{lk} = \begin{cases} 0 &\mbox{if } l < k \\ 
C_yB_u(k) & \mbox{if } l = k \\
C_yA(l)A(l-1)\cdots A(k)B_u(k) &\mbox{if } l > k \end{cases}
\end{equation} 
Where $C_y$ is the row of $C$ in (\ref{C4ad}) corresponding to the desired output and $B_u$ the column of $B$ in (\ref{C4ad})
corresponding to the desired input. Note that
for the case of uncoupled lateral and longitudinal dynamics, the off-diagonal sub-matrices of $P$ are $[0]$ since we have
two SISO systems. The term $\mathbf{w}$ in (\ref{eq:liftedDomain}) is the unknown disturbance. Iterative
learning control relies on the assumption that this disturbance is the underlying cause of the observed errors $\mathbf{e}_j$, and that the disturbance, while unknown, is constant from lap to lap.

>>>>

\begin{abstract}
Autonomous vehicles will generate data from the variety of sensors they employ to track the surrounding environment. This data is inherently valuable, as it gives algorithm designers the potential to leverage prior experience in order to improve driving performance over time. This paper uses the lens of autonomous racing to provide an example of how data from previous iterations of driving can be used to improve quantitative metrics of performance. Two complementary algorithms are demonstrated in this paper. The first algorithm uses iterative learning control (ILC) to simultaneously improve lateral and longitudinal tracking of the desired racing trajectory over multiple laps, while the second algorithm is focused on altering the trajectory itself using a search method. When driven experimentally at the limits of handling, the result is a reduction in lap time of nearly 1.4 seconds, a major improvement.

@@

\subsection{Problem Description and Proposed Solution}
Several methods exist in the literature to generate autonomous racing trajectories 
(\cite{casanova},\cite{kelly},\cite{theodosis},\cite{kapaniadscc}), which typically comprise of a longitudinal speed profile $U_x(s)$ and a desired curvature profile $\kappa(s)$ as a function of distance along the track. These trajectories can then be tracked using one of several path following controllers, such as (\cite{mickcop},\cite{fricStudy},\cite{mickgeneral},\cite{shladover},\cite{nagai}).

Using the trajectory generated by \cite{kapaniadscc} and the trajectory following controller presented by \cite{kapania}, the
authors generated a test run of an autonomous race vehicle (see Fig.\ref{fig:carpic}) at the limits of handling on the Thunderhill Raceway course in Willows, CA. %(Fig.~\ref{fig:carpic}(b)). 
The results are shown in Fig. \ref{fig:expErrors}.

\begin{figure}[h]
\centering
\includegraphics[width=2.5in]{figures/pic1.png}
\caption{``Shelley" - Stanford's autonomous self-driving race vehicle.}
\label{fig:carpic}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=2.8in]{figures/expErrors.eps}
\caption{Controller tracking performance and tire slip norm on a test run at the limits of handling ($\mu =$ 0.94).(a) Desired
vs actual speed of the vehicle. (b) Lateral tracking error and (c) tire slip norm as a function of distance along the path.}
\label{fig:expErrors}
\end{figure}

@@

Where $\sigma$ and $\alpha$ are the maximum longitudinal and lateral tire slips over all four tires, and $\sigma_p$ and $\alpha_p$ are empirically
determined \textit{peak slip} values resulting in maximum longitudinal and lateral tire force generation. As a result, $\zeta < 1 $ corresponds to the tires having excess force generation capacity, while $\zeta > 1$ corresponds to tire saturation.

@@

Iterative learning control (ILC) algorithms can help an autonomous vehicle follow a desired trajectory more precisely over several laps of driving. However, ILC algorithms do not alter the desired trajectory itself, only the input signals that attempt to track the trajectory. In cases where the trajectory results in oversteer or understeer, a trajectory modification algorithm is necessary since the vehicle may already be saturating the tires, and applying additional actuator effort will not result in better tracking.

Some trajectory modification algorithms previously investigated in
the literature have focused on modifying the \textit{lateral} trajectory by altering the curvature profile as the vehicle understeers or oversteers. For example, Theodosis and Gerdes presented an algorithm to gradually widen the radius of a turn in response to a detected understeer \cite{paulthesis}. Klomp and Gordon also developed a strategy for recovering from
vehicle understeer by solving for an optimal emergency braking profile to minimize deviation from the desired path \cite{klomp} Brown et al.~\cite{brown2017safe} presented a model predictive control (MPC) approach that generally aimed to follow a planned vehicle trajectory at the limits. However, if the vehicle was at risk of understeering or oversteering, the MPC algorithm would deviate laterally from the planned trajectory in order to maintain stability of the vehicle without driving off the road. Another approach to trajectory modification is Learning Model Predictive Control (LMPC), developed by Rosolia et al. \cite{rosolia2017autonomous} for the purpose of autonomous racing. In LMPC, data from every lap of racing is used to build invariant sets and approximate the value function for the generally non-convex lap time minimization problem.

@@

Due to the repetitive nature of automotive racing, iterative learning control techniques are a promising method to gradually eliminate \textit{trajectory tracking} errors. This paper will first present adaptations of two established ILC methods  
(proportional-derivative and quadratically optimal) for use in the Audi TTS racing system.  The ILC algorithms will be designed for combined lateral and longitudinal control, as an extension of the lateral-only ILC first presented in \cite{kapaniaacc}

Second, this paper will address situations where the vehicle underperforms due to excessive oversteer or understeer. An A* search algorithm is presented
that iteratively modifies portions of the \textit{velocity} profile to be more conservative if a stability violation is encountered on a prior lap. The algorithm also modifies the velocity profile to be more aggressive if the tires are not being actuated at the limits. In this way, the algorithm is focused on learning the time-optimal friction profile $\mu^\star(s)$ by searching through datasets obtained over multiple laps.

@@

The more difficult task is expressing the dynamics of the two subsystems using an affine model, given the tendency for the vehicle tires to saturate at the limits. This paper will model the dynamics with a bicycle model with single longitudinal input $F_x$ and front steer input $\delta$, shown in Fig.~\ref{fig:bikemodel}.

@@

% Nonlinear, coupled equations of motion are provided in (\ref{eq:fullNL})-(\ref{eq:fullNL2}):

% \begin{align}
% \label{eq:fullNL}
% 	\dfrac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
% 	\dfrac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{F_x}{m} - \frac{F_\mathrm{yf}(\alpha_\mathrm{f},F_x)\delta}{m}\\
% 	\dfrac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)}-r \\
% 	\dfrac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
% 	\dfrac{d\Delta\Psi}{dt} &= r - \kappa(s)(v+U^\mathrm{des}_x(s)) \label{eq:fullNL2}
% \end{align}
% The system dynamics presented in (\ref{eq:fullNL}) have five states, yaw rate $r$, vehicle sideslip $\beta$, lateral path tracking error $e$, heading error $\Delta \Psi$, and speed tracking error $v$, defined by:

% \begin{equation}
% v = U_x - U^\mathrm{des}_x
% \end{equation}

The potential for coupling between the subsystems occurs from both the state equations and also the complex nature of tire force generation at the handling limits. As shown in Fig.~\ref{fig:coupledTires}, as longitudinal force $F_x$ is distributed across the tires, the available lateral force decreases. At the limits of handling, this \textit{derating} of the lateral force may become significant. As a result, we model lateral tire force generation $F_y$ as a function of both lateral tire slip $\alpha$ and the applied longitudinal force $F_x$, using a modified form of the Fiala tire model \cite{rami}. 
\begin{figure}[tb]
\centering
\includegraphics[width=2.8in]{figures/tirecurves.eps}
\caption{Lateral tire force curve as a function of longitudinal force $F_x$ and lateral tire slip $\alpha$.}
\label{fig:coupledTires}
\end{figure}

The control inputs $F_x$ and $\delta$ are broken up into the closed-loop feedback term and the \textit{learned} component that is modified on every lap by the ILC algorithm:
\begin{align}
\label{eqn:splitTerms}
F_x &= F^{\mathrm{FB}}_x + F^L_x\\
    &= -K_xv + F^L_x\\
\delta &= \delta_{\mathrm{FB}} + \delta_L\\
\delta &= -k_p(e + x_\mathrm{LA}\Delta\Psi) + \delta_L
\end{align}
where $K_x$ is the proportional speed tracking gain and $k_p$ and $x_\mathrm{LA}$ are the lookahead gains presented in \cite{kapania}. The term $v$ denotes the speed tracking error.

The closed loop system dynamics are given by incorporating (\ref{eqn:splitTerms}) into equations of motion for the bicycle model: 
\begin{align}
\label{eq:fullNLCL}	
	\frac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
	\frac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{-K_xv + F^L_x}{m} \\
	& - \frac{F_\mathrm{yf}(\alpha_F,F_x)(k_p(e + x_{LA}\Delta\Psi) + \delta_L)}{m}\\
	\frac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)} - r \\
	\frac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
	\frac{d\Delta\Psi}{dt} &= r - \kappa(s)(v+U^\mathrm{des}_x(s))\label{eq:fullNLCL2}	
\end{align}

The nonlinear closed-loop dynamics must be converted into an affine, discrete-time dynamical system to apply conventional iterative learning
control algorithms. In our case, we have two system outputs ($y = [e \hspace{2mm} v]^T$) that are measured and two input signals to learn ($u = [\delta_L \hspace{2mm} F^L_x]^T$). 
Since we run the iterative learning control algorithm after seeing a trial of data, we can approximate
the dynamics in (\ref{eq:fullNLCL})-(\ref{eq:fullNLCL2}) by linearizing about the observed states and inputs $x_o, u_o$ from the first lap. These state estimates can be obtained from a differential GPS, simulation, or a state observer. The affine model is therefore given by: 
 \begin{align}
 \label{eq:ssMAT}
 \begin{bmatrix} \dot{e} \\ \dot{\beta} \\ \dot{r} \\ \dot{\Delta\Psi} \\ \dot{v}\end{bmatrix} &= \begin{bmatrix} A(t) \end{bmatrix} \begin{bmatrix} e \\ \beta \\ r \\ \Delta\Psi \\ v \end{bmatrix} + \begin{bmatrix} B(t) \end{bmatrix} \begin{bmatrix} \delta_L \\ F^L_x \end{bmatrix} + d(t) \\
 \begin{bmatrix} e \\ v \end{bmatrix}  &= \begin{bmatrix} 1 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 1 \end{bmatrix} x
 \end{align}

@@

 This multiple-input, multiple-output (MIMO) model captures the coupled behavior of the longitudinal and lateral inputs. If we make the assumption that the longitudinal and lateral inputs are decoupled and make small angle approximations, we may obtain the following resulting analytical state matrices $A(t)$ and $B(t)$ for (\ref{eq:ssMAT}):

\begin{align}
\label{eqn:C4dyn}
 &A(t)  = \\ \nonumber
  &\left[\begin{smallmatrix}
  0 & U_x(t) & 0 & U_x(t) & 0\\ 
  0 & 0 & 1 & 0 & 0 \\ 
  \frac{-ak_\mathrm{p} \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}}  & \frac{-ak_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}}  & \frac{-a^2\tilde{C}_\mathrm{f}(t)-b^2\tilde{C}_\mathrm{r}(t)}{U_x(t)I_\mathrm{z}} & \frac{b\tilde{C}_\mathrm{r}(t) - a\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 \\
  \frac{-k_\mathrm{p}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{-k_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{b\tilde{C}_\mathrm{r}(t)-a\tilde{C}_\mathrm{f}(t)}{mU_x(t)^2}-1 & \frac{-\tilde{C}_\mathrm{f}(t) - \tilde{C}_\mathrm{r}(t)}{mU_x(t)} & 0 \\
  0 & 0 & 0 & 0 & -\frac{K_x}{m}
  \end{smallmatrix}\right] \\
&B(t) =\begin{bmatrix} 0 & 0 \\	 	0 & 0 		\\ 		\frac{a \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 		\\ 		\frac{\tilde{C}_\mathrm{f}(t)}{mU_x(t)} & 0 		\\ 0 & \frac{1}{m} \end{bmatrix} 
\end{align}
The affine term $d(t)$ is given by: 
\begin{align}
d(t) = \left[\begin{matrix} 0 \\
               -\kappa(t) U_x(t) \\ 
			    \frac{a\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) - b\tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + a\tilde{F}_\mathrm{yf}(t) - b\tilde{F}_\mathrm{yr}(t)}{I_z}\\
				\frac{\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) + \tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + \tilde{F}_\mathrm{yf}(t) + \tilde{F}_\mathrm{yr}(t)}{mU_x(t)}\\
				0 \\
				0
				\end{matrix}\right]
\end{align}

@@

While beyond the scope of this paper, it should also be noted that advances in ILC research have focused specifically on ILC for nonlinear \cite{xunonlinear} and linear parameter-varying (LPV) systems \cite{de2017iterative}.

\subsection{Lifted Domain Representation and ILC \newline Problem Statement}
\label{sec:liftedD}
Whether the coupled (\ref{eq:C4nl}) or decoupled (\ref{eqn:C4dyn}) dynamics are assumed, the
final modeling step is to apply the matrix exponential discretization technique to obtain dynamics in the following form:
\begin{align}
\label{C4ad}
x_{k+1} &= A_kx_k + B_ku_k + d_k \\
y_{k}   &= Cx_k
\end{align}
For a given lap of racing $j$, sensor measurements provide $N$ observations of both the lateral path deviation $e$ and longitudinal
speed tracking error $v$. These measurements can be stacked into a 2$N \times 1$ array: 
\begin{equation}
\mathbf{e}_j = \begin{bmatrix} e_1 & \hdots & e_N & v_1 & \hdots & v_N \end{bmatrix}^T
\end{equation} 
These measurement errors are related to the learned control inputs $\delta^L$ and $F^L_x$ as follows: 
\begin{align}
\label{eq:liftedDomain}
\mathbf{e}_j &= P\mathbf{u}^L_j + \mathbf{w} \\
\mathbf{u}^L_j &= \begin{bmatrix} \delta^L_1 & \hdots & \delta^L_N & F^L_{x1} & \hdots & F^L_{xN} \end{bmatrix}^T
\end{align}
The system dynamics modeled in the previous section are represented by the \textit{lifted-domain} dynamics matrix $P$,
which is $\mathrm{2}N \times \mathrm{2}N$ and given by:
\begin{equation}
\label{eq:C4bm}
P=\left[
\begin{array}{c|c}
P_{e\delta} & P_{eF} \\ \hline
P_{v\delta} & P_{vF} 
\end{array}\right]
\end{equation}
Where each submatrix in (\ref{eq:C4bm}) is $N \times N$ and represents the lifted-domain dynamics from a given input to a given output. 
Individual terms of the sub-matrices are given by:

\begin{equation}
\label{eq:how2Lift}
p_{lk} = \begin{cases} 0 &\mbox{if } l < k \\ 
C_yB_u(k) & \mbox{if } l = k \\
C_yA(l)A(l-1)\cdots A(k)B_u(k) &\mbox{if } l > k \end{cases}
\end{equation} 
Where $C_y$ is the row of $C$ in (\ref{C4ad}) corresponding to the desired output and $B_u$ the column of $B$ in (\ref{C4ad})
corresponding to the desired input. The term $\mathbf{w}$ in (\ref{eq:liftedDomain}) is the unknown disturbance. Iterative
learning control relies on the assumption that this disturbance is the underlying cause of the observed errors $\mathbf{e}_j$, and that the disturbance, while unknown, is constant from lap to lap.
